Project Setup

### 2025-02-08 data.qmd
Reformat markdown
To ease change tracking in the future.
Too bad it wasn't done from the
beginning.

The data set is from the package [`palmerpenguins`](https://allisonhorst.github.io/palmerpenguins/) (v`r packageVersion("palmerpenguins")`) and contains the recorded bill lengths and sex of penguins living on three islands in the Palmer Archipelago, Antarctica. It was made available by Allison Horst, Alison Hill, and Kristen Gorman under the license [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/).

@@

Privacy laws and contractual obligations may require you to create a completely anonymized or synthetic data set^[For example, using [Amnesia](https://amnesia.openaire.eu/), [ARX](https://arx.deidentifier.org/), [sdcTools](https://github.com/sdcTools), or [Synthpop](https://www.synthpop.org.uk/).] (if possible), or prohibit any sharing of data, in which case you should provide a reference to a data repository where they can be obtained from. For further information, you can watch the talk "[Data anonymity](https://osf.io/j6fy8)" by Felix Schönbrodt recorded during the LMU Open Science Center Summer School 2023 and have a look at [the accompanying slides](https://osf.io/z6gcu). Another resource to look at is the presentation "Creating reproducible packages when data are confidential" by @Vilhuber2024.
:::

@@

When distributing a data set, it is important to document the meaning (e.g., units) and valid values of its variables. This is typically done with a _data dictionary_ (also called a _codebook_). In the following, we will demonstrate how to create a simple data dictionary using the R package [`pointblank`](https://rstudio.github.io/pointblank/). You can install it now using:

@@

Without the package `palmerpenguins` installed, you can find a [suggested citation on its website](https://allisonhorst.github.io/palmerpenguins/#citation) and run the following:

@@

For a more elaborate introduction to `pointblank`, you can read its "[Intro to Information Management](https://rstudio.github.io/pointblank/articles/INFO-1.html)".

One could go even further by making the information machine-readable in a standardized way. We provide an optional example of that in @nte-frictionless. If you want to learn more about the sharing of research data, have a look at the tutorial "[FAIR research data management](https://lmu-osc.github.io/FAIR-Data-Management/)".

@@

Specifically, one can use the R package [`frictionless`](https://docs.ropensci.org/frictionless/) to create a _schema_ which describes the structure of the data.^[In June 2024, [version 2](https://datapackage.org/) of the frictionless data standard has been released. As of November 2024, the R package `frictionless` only supports the first version, though support for v2 is [planned](https://github.com/frictionlessdata/frictionless-r/labels/datapackage%3Av2).] For the purpose of the following code, it is just a nested list that we edit to include our own information. We also explicitly record in the schema that missing values are stored in the data file as `NA` and that the data are licensed under CC0\ 1.0. Finally, the package is used to create a metadata file that contains the schema.

@@

All data relied upon should be cited in the manuscript to allow for precise identification and access. From the "eight core principles of data citation" by @Starr2015, licensed under [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/):

> **Principle 1 – Importance**: "Data should be considered legitimate, citable products of research. Data citations should be accorded the same importance in the scholarly record as citations of other research objects, such as publications."
> 
> **Principle 3 – Evidence**: "In scholarly literature, whenever and wherever a claim relies upon data, the corresponding data should be cited."
> 
> **Principle 5 – Access**: "Data citations should facilitate access to the data themselves and to such associated metadata, documentation, code, and other materials, as are necessary for both humans and machines to make informed use of the referenced data."
> 
> **Principle 7 – Specificity and Verifiability**: "Data citations should facilitate identification of, access to, and verification of the specific data that support a claim. Citations or citation metadata should include information about provenance and fixity sufficient to facilitate verifying that the specific time slice, version and/or granular portion of data retrieved subsequently is the same as was originally cited."

@@

You can find an appropriate BibTeX entry on [the package website](https://allisonhorst.github.io/palmerpenguins/#citation) or with the function `citation()`:^[Note that this function requires to have the respective package installed.]

>>>>

The data set is from the package [`palmerpenguins`][palmerpenguins]
(v`r packageVersion("palmerpenguins")`)
and contains the recorded bill lengths and sex of penguins
living on three islands in the Palmer Archipelago, Antarctica.
It was made available by Allison Horst, Alison Hill, and Kristen Gorman
under the license [CC0\ 1.0][cc0].

[palmerpenguins]: https://allisonhorst.github.io/palmerpenguins/
[cc0]: https://creativecommons.org/publicdomain/zero/1.0/

@@

Privacy laws and contractual obligations may require you
to create a completely anonymized or synthetic data set[^anon-synth-tools] (if possible),
or prohibit any sharing of data, in which case you should provide a reference
to a data repository where they can be obtained from.
For further information, you can watch the talk "[Data anonymity](https://osf.io/j6fy8)"
by Felix Schönbrodt recorded during the LMU Open Science Center Summer School 2023
and have a look at [the accompanying slides](https://osf.io/z6gcu).
Another resource to look at is the presentation
"Creating reproducible packages when data are confidential" by @Vilhuber2024.
:::

[^anon-synth-tools]: For example, using [Amnesia](https://amnesia.openaire.eu/),
[ARX](https://arx.deidentifier.org/), [sdcTools](https://github.com/sdcTools),
or [Synthpop](https://www.synthpop.org.uk/).

@@

When distributing a data set, it is important to document the meaning (e.g., units)
and valid values of its variables.
This is typically done with a _data dictionary_ (also called a _codebook_).
In the following, we will demonstrate how to create a simple data dictionary
using the R package [`pointblank`][pointblank].
You can install it now using:

[pointblank]: https://rstudio.github.io/pointblank/

@@

Without the package `palmerpenguins` installed,
you can find a [suggested citation on its website][palmerpenguins-citation]
and run the following:

[palmerpenguins-citation]: https://allisonhorst.github.io/palmerpenguins/#citation

@@

For a more elaborate introduction to `pointblank`,
you can read its "[Intro to Information Management][pointblank-intro]".

[pointblank-intro]: https://rstudio.github.io/pointblank/articles/INFO-1.html

One could go even further by making the information machine-readable in a standardized way.
We provide an optional example of that in @nte-frictionless.
If you want to learn more about the sharing of research data,
have a look at the tutorial "[FAIR research data management][fair-tutorial]".

[fair-tutorial]: https://lmu-osc.github.io/FAIR-Data-Management/

@@

Specifically, one can use the R package [`frictionless`][frictionless]
to create a _schema_ which describes the structure of the data.[^frictionless-note]
For the purpose of the following code,
it is just a nested list that we edit to include our own information.
We also explicitly record in the schema
that missing values are stored in the data file as `NA`
and that the data are licensed under CC0\ 1.0.
Finally, the package is used to create a metadata file that contains the schema.

[frictionless]: https://docs.ropensci.org/frictionless/

[^frictionless-note]: In June 2024, [version 2](https://datapackage.org/)
of the frictionless data standard has been released.
As of November 2024, the R package `frictionless` only supports the first version,
though support for v2 is [planned](https://github.com/frictionlessdata/frictionless-r/labels/datapackage%3Av2).

@@

All data relied upon should be cited in the manuscript
to allow for precise identification and access.
From the "eight core principles of data citation" by @Starr2015,
licensed under [CC0\ 1.0][cc0]:

> **Principle 1 – Importance**: "Data should be considered legitimate,
> citable products of research. Data citations should be accorded the same
> importance in the scholarly record as citations of other research objects,
> such as publications."
> 
> **Principle 3 – Evidence**: "In scholarly literature, whenever and wherever a
> claim relies upon data, the corresponding data should be cited."
> 
> **Principle 5 – Access**: "Data citations should facilitate access to the data
> themselves and to such associated metadata, documentation, code, and other
> materials, as are necessary for both humans and machines to make informed use
> of the referenced data."
> 
> **Principle 7 – Specificity and Verifiability**: "Data citations should
> facilitate identification of, access to, and verification of the specific data
> that support a claim. Citations or citation metadata should include
> information about provenance and fixity sufficient to facilitate verifying
> that the specific time slice, version and/or granular portion of data
> retrieved subsequently is the same as was originally cited."

@@

You can find an appropriate BibTeX entry on [the package website][palmerpenguins-citation]
or with the function `citation()`:^[Note that this function requires
to have the respective package installed.]

### 2025-02-08 intro.qmd
Reformat markdown
To ease change tracking in the future.
Too bad it wasn't done from the
beginning.

Suppose you are reading an article about a new imaging method to turn seismological data into subsurface images. The article describes the ideas that went into developing this method and presents a few examples to illustrate its superiority over previous approaches. You got interested and would like to apply this method to your own data. However, with only the article available, it could take months to come up with a working solution, if possible at all. This situation has been put aptly by Buckheit & Donoho [-@Buckheit1995, p. 59], distilling an idea by the geophysicist Jon Claerbout:

> "An article about computational science in a scientific publication is __not__ the scholarship itself, it is merely __advertising__ of the scholarship. The actual scholarship is the complete software development environment and the complete set of instructions which generated the figures."

@@

The practice to interleave narrative text with code has its roots in the paradigm of literate programming, where documentation and source code are treated as equals and are arranged in a way to maximize understanding [@Knuth1984]. Alternating text and code can be also found in notebook interfaces for exploratory programming [e.g., Wolfram Mathematica or Jupyter Notebooks, see @Kluyver2016] that also have the capability to execute code and embed its output. With _Sweave_ [@Leisch2002], ideas from both worlds -- literate programming and embedding program output -- were combined into one tool for rendering dynamic documents using the R programming language. It is the predecessor of the R package `knitr` [@Xie2015] which is being used under the hood in this tutorial.^[Specifically, Quarto employs `knitr` to execute chunks of R code.]

@@

> __Does the research compendium contain everything needed to reproduce a predefined outcome in an organized and parsimonious way?__
> 
> 1. __Completeness:__ The research compendium contains all of the objects needed to reproduce a predefined outcome.
> 2. __Organization:__ It is easy to understand and keep track of the various objects in the research compendium and their relationship over time.
> 3. __Economy:__ Fewer extraneous objects in the compendium mean fewer things that can break and require less maintenance over time.
> 
> __Is descriptive information about the research compendium and its components available and easy to understand?__
> 
> 4. __Transparency:__ The research compendium provides full disclosure of the research process that produced the scientific claim.
> 5. __Documentation:__ Information describing compendium objects is provided in enough detail to enable independent understanding and use of the compendium.
> 
> __Is information about how the research compendium and its components can be used available and easy to understand?__
> 
> 6. __Access:__ It is clear who can use what, how, and under what conditions, with open access preferred.
> 7. __Provenance:__ The origin of the components of the research compendium and how each has changed over time is evident.
> 
> __Is information about the research compendium and its components embedded in code?__
> 
> 8. __Metadata:__ Information about the research compendium and its components is embedded in a standardized, machine-readable code.
> 9. __Automation:__ As much as possible, the computational workflow is script- or workflow-based so that the workflow can be re-executed using minimal actions.
> 
> __Is there a plan for reviewing the research compendium for FAIR and computational reproducibility standards over time?__
> 
> 10. __Review:__ A series of managed activities are needed to ensure continued access to and functionality of the research compendium and its components for as long as necessary.

>>>>

Suppose you are reading an article about a new imaging method
to turn seismological data into subsurface images.
The article describes the ideas that went into developing this method
and presents a few examples to illustrate its superiority over previous approaches.
You got interested and would like to apply this method to your own data.
However, with only the article available, it could take months
to come up with a working solution, if possible at all.
This situation has been put aptly by
Buckheit & Donoho [-@Buckheit1995, p. 59, emphasis in original],
distilling an idea by the geophysicist Jon Claerbout:

> "An article about computational science in a scientific publication
> is __not__ the scholarship itself, it is merely __advertising__ of the scholarship.
The actual scholarship is the complete software development environment
and the complete set of instructions which generated the figures."

@@

The practice to interleave narrative text with code has its roots
in the paradigm of literate programming, where documentation and source code
are treated as equals and are arranged in a way to maximize understanding [@Knuth1984].
Alternating text and code can be also found in notebook interfaces for exploratory programming
[e.g., Wolfram Mathematica or Jupyter Notebooks, see @Kluyver2016]
that also have the capability to execute code and embed its output.
With _Sweave_ [@Leisch2002], ideas from both world
-- literate programming and embedding program output --
were combined into one tool for rendering dynamic documents using the R programming language.
It is the predecessor of the R package `knitr` [@Xie2015]
which is being used under the hood in this tutorial.[^knitr]

[^knitr]: Specifically, Quarto employs `knitr` to execute chunks of R code.

@@

> __Does the research compendium contain everything needed to reproduce a
> predefined outcome in an organized and parsimonious way?__
> 
> 1. __Completeness:__ The research compendium contains all of the objects
>    needed to reproduce a predefined outcome.
> 2. __Organization:__ It is easy to understand and keep track of the various
>    objects in the research compendium and their relationship over time.
> 3. __Economy:__ Fewer extraneous objects in the compendium mean fewer things
>    that can break and require less maintenance over time.
> 
> __Is descriptive information about the research compendium and its components
> available and easy to understand?__
> 
> 4. __Transparency:__ The research compendium provides full disclosure of
>    the research process that produced the scientific claim.
> 5. __Documentation:__ Information describing compendium objects is provided
>    in enough detail to enable independent understanding and use of the compendium.
> 
> __Is information about how the research compendium and its components
> can be used available and easy to understand?__
> 
> 6. __Access:__ It is clear who can use what, how, and under what conditions,
>    with open access preferred.
> 7. __Provenance:__ The origin of the components of the research compendium
>    and how each has changed over time is evident.
> 
> __Is information about the research compendium and its components embedded in code?__
> 
> 8. __Metadata:__ Information about the research compendium and its component
>    is embedded in a standardized, machine-readable code.
> 9. __Automation:__ As much as possible, the computational workflow is script-
>    or workflow-based so that the workflow can be re-executed using minimal actions.
> 
> __Is there a plan for reviewing the research compendium for FAIR
> and computational reproducibility standards over time?__
> 
> 10. __Review:__ A series of managed activities are needed to ensure
>    continued access to and functionality of the research compendium and its components for as long as necessary.

### 2025-02-08 make_readme.qmd
Reformat markdown
To ease change tracking in the future.
Too bad it wasn't done from the
beginning.

:   Badges typically report quick facts, like [whether the project is actively maintained](https://www.repostatus.org/), [how many dependencies it has](https://tinyverse.netlify.app/), or [whether it has been published in a package repository](https://www.r-pkg.org/services#badges). Sometimes, the license, any associated DOIs, or quality metrics like code coverage by unit tests are communicated via badges as well.

@@

:   What steps need to be taken to run the project? What software needs to be installed? R itself and the R packages are already documented as this project uses `renv`. Therefore you can focus on all other dependencies, such as the system dependencies of R packages as well as the version of Quarto.^[As of August 2024, a proposal for `renv` to record the version of Quarto has not been implemented, see [rstudio/renv#1143](https://github.com/rstudio/renv/issues/1143).] Also, don't forget to mention software that you have used for any manual steps. See @sec-dependencies for additional information. It may make sense to provide these information in the README even if you already cite all programs and their version numbers in your manuscript.

@@

If you installed `apaquarto` or any other Quarto extension, one can query their versions as follows:^[Luckily, the extensions are included in the project folder, so technically their version is already recorded in the project's files.]

>>>>

:   Badges typically report quick facts, like
    [whether the project is actively maintained][repostatus],
    [how many dependencies it has][tinyverse],
    or [whether it has been published in a package repository][metacran].
    Sometimes, the license, any associated DOIs,
    or quality metrics like code coverage by unit tests
    are communicated via badges as well.

[repostatus]: https://www.repostatus.org/
[tinyverse]: https://tinyverse.netlify.app/
[metacran]: https://www.r-pkg.org/services#badges

@@

:   What steps need to be taken to run the project? What software needs to be installed?
    R itself and the R packages are already documented as this project uses `renv`.
    Therefore you can focus on all other dependencies,
    such as the system dependencies of R packages
    as well as the version of Quarto.[^renv-quarto]
    Also, don't forget to mention software that you have used for any manual steps.
    See @sec-dependencies for additional information.
    It may make sense to provide these information in the README
    even if you already cite all programs and their version numbers in your manuscript.

[^renv-quarto]: As of August 2024, a proposal for `renv` to record the version of Quarto
has not been implemented, see [rstudio/renv#1143](https://github.com/rstudio/renv/issues/1143).

@@

If you installed `apaquarto` or any other Quarto extension,
one can query their versions as follows:[^already-included]

[^already-included]: Luckily, the extensions are included in the project folder,
so technically their version is already recorded in the project's files.

### 2025-02-08 setup.qmd
Reformat markdown
To ease change tracking in the future.
Too bad it wasn't done from the
beginning.

As we will use Git to track the version history of files, be sure to check _Create a git repository_. If you don't know what Git is, have a look at the tutorial "[Introduction to version control with git and GitHub within RStudio](https://lmu-osc.github.io/Introduction-RStudio-Git-GitHub/)".

@@

Also, we will utilize the package [`renv`](https://rstudio.github.io/renv/) to track the R packages our project depends on. Using it makes it easier for others to view and obtain them at the exact same version at a later point in time. Therefore make sure that the box _Use renv with this project_ is checked. Again, if this is the first time you are hearing about `renv`, have a look at the tutorial "[Introduction to {renv}](https://lmu-osc.github.io/introduction-to-renv/)".

>>>>

As we will use Git to track the version history of files,
be sure to check _Create a git repository_.
If you don't know what Git is, have a look at the tutorial
"[Introduction to version control with git and GitHub within RStudio][git-tutorial]".

[git-tutorial]: https://lmu-osc.github.io/Introduction-RStudio-Git-GitHub/

@@

Also, we will utilize the package [`renv`][renv]
to track the R packages our project depends on.
Using it makes it easier for others to view and obtain them
at the exact same version at a later point in time.
Therefore make sure that the box _Use renv with this project_ is checked.
Again, if this is the first time you are hearing about `renv`,
have a look at the tutorial "[Introduction to {renv}][renv-tutorial]".

[renv]: https://rstudio.github.io/renv/
[renv-tutorial]: https://lmu-osc.github.io/introduction-to-renv/
